A Multiple-Set Overlapping-Domain Decomposed Monte Carlo Synthetic Acceleration Method for Linear Systems

### 2013-09-08 doc/sna2013/paper/slattery_evans_wilson.tex
finished sna draft

\abstract{We present a multiple-set overlapping-domain decomposed
  strategy for parallelizing the Monte Carlo Synthetic Acceleration
  method. Monte Carlo Synthetic Acceleration methods use the
  Neumann-Ulam class of Monte Carlo solvers for linear systems to
  accelerate a fixed-point iteration sequence. Effective parallel
  algorithms for these methods require the parallelization of the
  underlying Neumann-Ulam solvers. To do this in a domain decomposed
  environment, we borrow strategies traditionally implemented in Monte
  Carlo particle transport to parallelize the problem. The parallel
  Neumann-Ulam and multiple-set overlapping-domain decomposition
  algorithm is presented along with parallel scaling data for the
  resulting implementation using the Titan Cray XK7 machine at Oak
  Ridge National Laboratory.}

@@

Finally, the expectation value using all permutations is:

@@

\subsection{Monte Carlo Synthetic Acceleration}
Using the ideas of Halton's residual method
\cite{halton_sequential_1994}, Evans and Mosher recently developed a
Monte Carlo solution method that was not prohibited severely by the
quality of the initial guess for the system \cite{evans_monte_2009}
and later applied it more rigorously as a solution mechanism for the
radiation diffusion equation \cite{evans_monte_2012}. With their new
methods, they achieved identical numerical results when compared to
conventional subspace methods and observed better performance with
respect to time to solution. Their approach was instead to use
residual Monte Carlo as a synthetic acceleration for a stationary
method.

@@

\begin{figure}[h!]
  \begin{center}
    \scalebox{0.7}{ \input{domain_to_domain.pdftex_t} }
  \end{center}
  \caption{Nearest neighbor history communication sequence.
    \textit{Each subdomain in the system has a set of nearest
      neighbors determined by the parallel adjacency graph of the
      input matrix. The subdomains are indexed by an integer.}}
  \label{fig:nearest_neighbor_comm}
\end{figure}
Given the mutual exclusivity of the transport algorithm, an additional
communcation sequence is required to efficiently stop the algorithm
after all histories have been computed. To stop transport, Brunner and
Brantley use an asynchronous binary communication tree as presented
for the same 9 subdomain example in
Figure~\ref{fig:binary_comm_tree}. In this communication pattern, each
process has one parent process (except for process 0) to which it will
non-blocking send the number of histories that terminated their
transport procedure by weight cutoff in the local
domain. Equivalently, each process has up to two child processes from
which it will receive their completed number of histories with a
non-blocking receive operation. Setting up a tree in this manner lets
the completed history tally be updated incrementally and funneled to
the root process in a mutually exclusive manner. Once the root process
has tallied a completed number of histories equal to the number of
histories in the source, transport is complete. Once this occurs, the
root process non-blocking sends the stop message to its children and
each child process non-blocking receives a stop message from the root
as shown in Figure~\ref{fig:binary_comm_tree}. The stop message is
then propagated up the tree in the same manner between parents and
children.
\begin{figure}[h!]
  \begin{center}
    \scalebox{0.6}{
      \input{binary_comm_tree.pdftex_t} }
  \end{center}
  \caption{Binary communication tree for coordinating the end of a
    parallel Neumann-Ulam solve. \textit{Each child process reports to
      its parents how many histories completed within its domain. When
      the root process sums all complete histories, it forwards the
      stop signal to its children which in turn forward the message up
      the tree. The subdomains are indexed by an integer.}}
  \label{fig:binary_comm_tree}
\end{figure}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Parallel Scaling Studies}
We will next measure performance of the parallel MCSA algorithm in
various scaling studies to assess its quality using a simple
two-dimensional neutron diffusion problem in a homogeneous medium. For
the scaling results presented here, all parallel MCSA results were
computed using the Monte Carlo Linear Solvers\footnote{MCLS is an open
  source C++ library built on the Trilinos\cite{heroux_overview_2005}
  framework and can be downloaded at
  \url{https://github.com/sslattery/MCLS}} (MCLS) library which
contains implementations of the parallel Neumann-Ulam and MCSA
algorithms. For every scaling study presented, the calculations were
performed 3 times with the average values reported for wall time and
those average values used for the efficiency computations. All
parallel scaling studies presented in this work were performed on the
Titan Cray XK7 machine at the Oak Ridge Leadership Computing Facility
located at Oak Ridge National Laboratory.

For all cases, both strong and weak scaling studies were
performed. For strong scaling, the objective is to decrease the
runtime of a fixed size global problem as a linear function of $P$,
where $P$ is the number of processors being used. Based on this, we
can define the strong scaling \textit{absolute efficiency} as
\cite{keyes_how_1999}:
\begin{equation}
  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
  \label{eq:strong_scaling_absolute}
\end{equation}
where $T(N,P)$ is the wall time for a computation of global size $N$
using $P$ processes. Note here that this particular definition is
measured with respect to a serial computation. On leadership-class
machines, it is often the case that we use a reference case that
occupies an entire node in the system. Therefore, we consider absolute
scaling for a reference case where $Q$ processes are used instead of a
serial computation:
\begin{equation}
  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
  \label{eq:strong_scaling_absolute_ref}
\end{equation}
For weak scaling, the ratio $N/P$ is fixed such that the local problem
size remains the same for all values of $P$. In this case, the
objective is to maintain a constant wall time while both $N$ and $P$
grow as a fixed ratio. In this case, we define the absolute efficiency
as \cite{keyes_how_1999}:
\begin{equation}
  \eta_{weak}(M|N,P|Q) = \frac{T(M,Q)}{T(N,P)}\:,
  \label{eq:weak_scaling_absolute}
\end{equation}
subject to the constraint $M/Q = N/P$. The scaling is perfect if the
runtime is static for all $P$ when $N/P$ is fixed.

@@

\subsection{Pure Domain Decomposition Scaling}
\label{subsec:pure_domain_decomp}
For the first set of scaling studies, we consider the parallel MCSA
algorithm where MSOD has not been used in the parallel Neumann-Ulam
algorithm. We refer to this case as pure domain decomposition where no
overlap or replication has been leveraged. Such a test will
specifically isolate the performance of Brunner and Brantley's
algorithm as applied to the random walk sequence. For the strong
scaling study, the global diffusion problem size was fixed at a
relatively small \sn{1.6}{7} DOFs and the 16 core run used as the
reference case. Figure~\ref{fig:titan_pure_strong_time} gives the wall
time per iteration for these calculations while
Figure~\ref{fig:titan_pure_strong} gives the absolute efficiencies
computed using Eq~(\ref{eq:strong_scaling_absolute}). All MCSA
calculations converged the diffusion problem in 22 iterations.

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_pure_strong_time.pdf}
  \end{center}
  \caption{Pure domain decomposition strong scaling wall time per
    iteration.}
  \label{fig:titan_pure_strong_time}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_pure_strong.pdf}
  \end{center}
  \caption{Pure domain decomposition strong scaling absolute
    efficiency. \textit{Super-linear speed-up is from memory thrashing
      in the reference case.}}
  \label{fig:titan_pure_strong}
\end{figure}

@@

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_pure_weak_time.pdf}
  \end{center}
  \caption{Pure domain decomposition weak scaling wall time
      per iteration.}
  \label{fig:titan_pure_weak_time}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_weak_absolute.pdf}
  \end{center}
  \caption{Pure domain decomposition weak scaling absolute
    efficiency. \textit{MCLS is an over order of magnitude slower
      arithmetically, causing the higher efficiency. GMRES executes
      more operations than conjugate gradient, also creating a higher
      efficiency.}}
  \label{fig:titan_weak_absolute}
\end{figure}

\subsection{Overlapping Domain Decomposition}
\label{subsec:overlapping_domain_decomp}
In the next set of scaling studies, we will explore how adding overlap
to the subdomains in the problem affects run times and parallel
performance as compared to the case with pure domain decomposition. To
do this, we utilize the set of analytic relations developed in
\cite{slattery_spectral_2013} to design these numerical
experiments. For the neutron diffusion problem used in these scaling
studies, the linear operator had a spectral radius of 0.787 and a
weight cutoff of \sn{1}{-2} was used with the Neumann-Ulam
solver. Using Eq~(\ref{eq:analytic_k}) gives an expected random walk
length of 19.22 transitions for any given history in the
problem. Using the mean-chord approximation given by
Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
5.3\% of the histories born in a particular domain to leak out after
the first transport stage. By adding overlap, we aim to reduce this
fraction of communicated histories and therefore boost the parallel
performance of the algorithm.

To determine the amount of overlap to test for the scaling studies, we
use a variation of Eq~(\ref{eq:step_k_length}) to determine how many
discrete states a history will move on average from its birth site by
modifying it to only account for discrete movements:
\begin{equation}
  \sqrt{\bar{n^2_k}} = n_s \sqrt{k}\:,
  \label{eq:discrete_distance}
\end{equation}
where $\sqrt{\bar{n^2_k}}$ is now the root-mean-squared number of
discrete states a history will move from its birth site. For our
diffusion problem used in these scaling studies, we compute a value of
2.63 states per history for any of the dimensions of the system. This
means that we can expect to end up on average 2.63 states in both the
$i$ and $j$ dimensions away from the starting state. Therefore, we
will use the overlap generation algorithm outlined in
\S~\ref{subsec:msod_generation} such that we parametrically grow the
local domain using 1, 2, 3, and 5 levels to measure how scalability is
affected when this root-mean-squared number of states is
considered. For example, using 3 levels of overlap should keep over
50\% of the histories that would be communicated in the pure domain
decomposition case on-process as this is greater than the
root-mean-squared number of states we would expect a history to
travel.

For the performance metrics, we may use the same weak and strong
scaling efficiency metrics that we used for the pure domain
decomposition case. As we are comparing to the case without overlap
and are looking to boost performance relative to this case, for both
weak and strong scaling calculations we will use the 16-core case
without overlap as the reference computation for all efficiency
measures reported. The absolute efficiency results from the strong
scaling exercise are given in
Figure~\ref{fig:titan_strong_overlap}. Aside from the boosting of
efficiencies at 2,048 cores, little evidence of strong scaling
improvement is evident for all levels of overlap tested.

@@

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_strong_overlap.pdf}
  \end{center}
  \caption{Strong scaling absolute efficiency with varying levels of
    overlap. \textit{Values of 0, 1, 2, 3, and 5 levels of overlap
      were used as the problem had a root-mean-squared number of
      states of 2.63 for each history.}}
  \label{fig:titan_strong_overlap}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3in]{titan_strong_overlap_diff.pdf}
  \end{center}
  \caption{Strong scaling absolute efficiency difference with
      respect to the 0 overlap reference case. \textit{Values of 1, 2, 3,
      and 5 levels of overlap are compared here with the absolute
      difference of the absolute efficiencies presented.}}
  \label{fig:titan_strong_overlap_diff}
\end{figure}

For weak scaling, the same study as in the pure domain decomposition
case was performed again with 1, 2, 3, and 5 levels of overlap. The
absolute parallel efficiencies from these calculations are given in
Figure~\ref{fig:titan_weak_overlap} and the differences in
efficiencies from the calculations without overlap given in
Figure~\ref{fig:titan_weak_overlap_diff}. The results here are much
the same in that there is not a dramatic improvement in scalability
relative to the case without overlap. Opposite the strong scaling
calculation, overlap enhances weak scaling efficiencies at lower core
counts and degrades the efficiencies at higher core counts. In
addition, 1 level of overlap provided the best enhancement of
scalability when overlap was beneficial.

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_weak_overlap.pdf}
  \end{center}
  \caption{Weak scaling absolute efficiency with varying levels of
    overlap. \textit{Values of 0, 1, 2, 3, and 5 levels of overlap
      were used as the problem had a root-mean-squared number of
      states of 2.63 for each history.}}
  \label{fig:titan_weak_overlap}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_weak_overlap_diff.pdf}
  \end{center}
  \caption{Weak scaling absolute efficiency difference with respect to
    the 0 overlap reference case. \textit{Values of 1, 2, 3, and 5
      levels of overlap are compared here with the absolute difference
      of the absolute efficiencies presented.}}
  \label{fig:titan_weak_overlap_diff}
\end{figure}

For both the strong and weak scaling studies, the parallel performance
of the algorithm was either marginally improved or not improved at
all. We would always expect an improvement in performance for using
overlap to prevent the transport of histories from domain to domain
during execution of the Neumann-Ulam algorithm. The reduction in
transport means a reduction in parallel communication and therefore
should never degrade the performance of the algorithm. It is true that
doing so in fact enhances Monte Carlo performance. However, the
Neumann-Ulam algorithm is not representative of the entire MCSA
algorithm. When overlap is used, an overlapping tally vector is
constructed in parallel such that multiple cores may own multiple
states in the system. When this occurs, the tally data associated with
these shared states must be communicated after the Neumann-Ulam solve
as outlined in \S~\ref{subsec:msod_algorithm}. This reduction
operation then builds the full MCSA correction vector in the
decomposition of the solution vector.

@@

\subsection{Multiple Set Domain Decomposition}
\label{subsec:ms_decomposition}
We next look to apply replication to the problem in the form of
multiple sets. As the previous analysis showed overlap to be of little
assistance in improving scalability, we only apply multiple sets to
the problem with pure domain decomposition within the sets for these
scaling studies. When adding sets, we must decide how many stochastic
histories to use to generate the correction vector in the MCSA
iteration. For this work we consider two choices; the split case and
the fully replicated case. As an example of these cases, consider a
single set problem where 20,000 histories are used in the Monte Carlo
solve. If we choose the split case, then solving the same problem with
2 sets would give 10,000 histories in each set and 4 sets would give
5,000 histories in each set. Splitting the histories in this way
provides the same global number of histories used to compute the
correction in the single set problem and we should therefore expect
the same iterative performance at a reduced time per iteration and
thus reduced time to solution. For the replicated history case, each
additional set in the problem will use just as many histories as the
single set case. Using 2 sets with fully replicated histories for our
example problem then has each set computing 20,000 histories with
40,000 global histories in the computation. For the 4 set case, the
number of global histories is increased to 80,000. By increasing the
global number of histories, the replicated cases should approximately
maintain time per iteration but should decrease the number of
iterations required to converge and subsequently reduce time to
solution.

\subsubsection{Strong Scaling with Multiple Sets}
\label{subsubsec:ms_strong}
For the strong scaling case, recall that the objective is to decrease
the time to solution for a given global problem size $N$ by increasing
$P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref}),
(\ref{eq:algorithmic_efficiency}), and
(\ref{eq:implementation_efficiency}) to compute the various strong
scaling efficiencies, but we must modify the reference computation in
each of these formulas to reflect our objective. For all efficiency
computations, the 16 core single set computation will be used as the
reference case. By adding more cores to the problem, either by
replicating with multiple sets, or decreasing the local problem size,
any computational resources added to the problem should aim to result
in speedup of this reference case. If this speedup for one technique
of adding cores is less than another, then that technique is deemed a
less efficient use of those computational resources.

For the multiple sets problem, the single set strong scaling study
from the pure domain decomposition reference case was repeated for
multiple sets with both splitting of histories across sets and full
replication of histories across sets using both 2 and 4
sets. Figure~\ref{fig:titan_strong_ms_time} gives the actual wall
times for these computations. There are two important features to note
in the wall time data. First, splitting the histories among sets
instead of replicating them is faster due to the fact that cutting the
number of histories in half reduces the time of an iteration by a
factor that is nearly linear while increasing linearly the number of
histories through replication does not decrease the iterations
required to converge by a linear factor. Second, until the strong
scaling wall is hit starting at around 10,000 cores, the single set
case actually performs the fastest with only marginal improvement in
run times at higher numbers of cores with more sets. At these larger
numbers of cores, however, the improvements are so modest that one is
better off running the same problem using a single set at lower core
counts to conserve computational resources while maintaining a good
time to solution.

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3in]{titan_strong_ms_time.pdf}
  \end{center}
  \caption{Wall time in seconds to solution for each case for the
    strong scaling study with multiple sets. \textit{Until the strong
      scaling wall is reached at O(10,000) cores, the single set case
      performs the best. For the multiple set cases, it is faster to
      split histories among sets and maintain iterative performance
      rather than replicate histories to improve iterative
      performance.}}
  \label{fig:titan_strong_ms_time}
\end{figure}

>>>>

\abstract{We present a multiple-set overlapping-domain decomposed
  strategy for parallelizing the Monte Carlo Synthetic Acceleration
  method. Monte Carlo Synthetic Acceleration methods use the
  Neumann-Ulam class of Monte Carlo solvers for linear systems to
  accelerate a fixed-point iteration sequence. Effective parallel
  algorithms for these methods require the parallelization of the
  underlying Neumann-Ulam solvers. To do this in a domain decomposed
  environment, we borrow strategies traditionally implemented in Monte
  Carlo particle transport to parallelize the problem. The parallel
  Neumann-Ulam and multiple-set overlapping-domain decomposition
  algorithms are presented along with parallel scaling data for the
  resulting implementation using the Titan Cray XK7 machine at Oak
  Ridge National Laboratory.}

@@

The expectation value using all permutations is:

@@

\subsection{Monte Carlo Synthetic Acceleration}
Using the ideas of Halton's residual method
\cite{halton_sequential_1994}, Evans and Mosher recently developed a
Monte Carlo solution method that was not prohibited severely by the
quality of the initial guess for the system \cite{evans_monte_2009}
and later applied it more rigorously as a solution mechanism for the
radiation diffusion equation \cite{evans_monte_2012}. With their new
methods, they achieved identical numerical results when compared to
conventional subspace methods (e.g. Krylov solvers) and observed
better performance with respect to time to solution. Their approach
was instead to use residual Monte Carlo as a synthetic acceleration
for a stationary method.

@@

\begin{figure}[h!]
  \begin{center}
    \scalebox{0.7}{ \input{domain_to_domain.pdftex_t} }
  \end{center}
  \caption{Nearest neighbor history communication sequence.
    \textit{Each subdomain in the system has a set of nearest
      neighbors determined by the parallel adjacency graph of the
      input matrix. The subdomains are indexed by an integer.}}
  \label{fig:nearest_neighbor_comm}
\end{figure}
Given the mutual exclusivity of the transport algorithm, an additional
communcation sequence is required to efficiently stop the algorithm
after all histories have been computed. To stop transport, Brunner and
Brantley use an asynchronous binary communication tree as presented
for the same 9 subdomain example in
Figure~\ref{fig:binary_comm_tree}. In this communication pattern, each
process has one parent process (except for process 0) to which it will
non-blocking send the number of histories that terminated their
transport procedure by weight cutoff in the local
domain. Equivalently, each process has up to two child processes from
which it will receive their completed number of histories with a
non-blocking receive operation. Setting up a tree in this manner lets
the completed history tally be updated incrementally and funneled to
the root process in a mutually exclusive manner. Once the root process
has tallied a completed number of histories equal to the number of
histories in the source, transport is complete. Once this occurs, the
root process non-blocking sends the stop message to its children and
each child process non-blocking receives a stop message from the root
as shown in Figure~\ref{fig:binary_comm_tree}. The stop message is
then propagated up the tree in the same manner between parents and
children.
\begin{figure}[h!]
  \begin{center}
    \scalebox{0.6}{
      \input{binary_comm_tree.pdftex_t} }
  \end{center}
  \caption{Binary communication tree for coordinating the end of a
    parallel Neumann-Ulam solve. \textit{Each child process reports to
      its parents how many histories completed within its domain. When
      the root process history sum equals the total number of
      histories, it forwards the stop signal to its children which in
      turn forward the message up the tree. The subdomains are indexed
      by an integer.}}
  \label{fig:binary_comm_tree}
\end{figure}

@@

To implement the MSOD algorithm, we physically copy all data
structures required to implement the Neumann-Ulam algorithm to each of
the sets in the problem. Within each set, the subdomains detect their
neighboring subdomains and gather the number of adjacent states
specified for overlap. For each of these overlap states, the
probablities, weights, tallies, and other information required to
implement the Monte Carlo sequence are gathered. When transport
begins, within each individual set the Neumann-Ulam sequence is
executed with parallelism provided Brunner and Brantely's algorithm as
outlined in the previous section. Here, transport occurs both in the
local domain and the overlap region before histories are
communicated. Once transport in all sets is compelete, the tallies
from all overlapping states are gathered to their original owning
processors within each set and then all tallies are summed onto the
set of processors occupied by the linear problem to construct the
final tally vector.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Parallel Scaling Studies}
We will next measure the performance of the parallel MCSA algorithm in
various scaling studies to assess its quality using a simple
two-dimensional neutron diffusion problem in a homogeneous medium. For
MCSA, a range of MSOD parameters will be used such that problems with
multiple sets may be tested along with overlap. For all MCSA
calculations, a single history was used for every degree of freedom in
the problem to compute the correction vector. For all calculations,
the flux was converged to a tolerance of
$\epsilon = \sn{1}{-8}$. Table~\ref{tab:parallel_mcsa_parameters} gives the
parameters for the diffusion problem.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lc}\hline\hline
      \multicolumn{1}{l}{Parameter}& 
      \multicolumn{1}{c}{Value}\\\hline
      $dx$ & 0.1 \\
      $dy$ & 0.1 \\
      $\Sigma_a$ & 5.0 \\
      $\Sigma_s$ & 1.0 \\
      Boundary Type & Vacuum \\
      MCSA Solver Tolerance & \sn{1}{-8} \\
      Histories & 1 per DOF \\
      Weight Cutoff & \sn{1}{-2} \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Parallel MCSA verification model problem parameters.}
    \textit{The neutron diffusion equation in 2 dimensions is used for
      the model problem.}}
  \label{tab:parallel_mcsa_parameters}
\end{table}
For the scaling results presented here, all parallel MCSA results were
computed using the Monte Carlo Linear Solvers\footnote{MCLS is an open
  source C++ library built on the Trilinos\cite{heroux_overview_2005}
  framework and can be downloaded at
  \url{https://github.com/sslattery/MCLS}} (MCLS) library which
contains implementations of the parallel Neumann-Ulam and MCSA
algorithms. For every scaling study presented, the calculations were
performed 3 times with the average values reported for wall time and
those average values used for the efficiency computations. All
parallel scaling studies presented in this work were performed on the
Titan Cray XK7 machine at the Oak Ridge Leadership Computing Facility
located at Oak Ridge National Laboratory.

For all cases, both strong and weak scaling studies were
performed. For strong scaling, the objective is to decrease the
runtime of a fixed size global problem as a linear function of $P$,
where $P$ is the number of processors being used. Based on this, we
define the strong scaling \textit{absolute efficiency} as
\cite{keyes_how_1999}:
\begin{equation}
  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
  \label{eq:strong_scaling_absolute}
\end{equation}
where $T(N,P)$ is the wall time for a computation of global size $N$
using $P$ processes. Note here that this particular definition is
measured with respect to a serial computation. On leadership-class
machines, it is often the case that we use a reference case that
occupies an entire node in the system. Therefore, we consider absolute
scaling for a reference case where $Q$ processes are used instead of a
serial computation:
\begin{equation}
  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
  \label{eq:strong_scaling_absolute_ref}
\end{equation}
For weak scaling, the ratio $N/P$ is fixed such that the local problem
size remains the same for all values of $P$. In this case, the
objective is to maintain a constant wall time while both $N$ and $P$
grow as a fixed ratio. In this case, we define the absolute efficiency
as \cite{keyes_how_1999}:
\begin{equation}
  \eta_{weak}(M|N,P|Q) = \frac{T(M,Q)}{T(N,P)}\:,
  \label{eq:weak_scaling_absolute}
\end{equation}
subject to the constraint $M/Q = N/P$. The scaling is perfect if the
runtime is static for all $P$ when $N/P$ is fixed.

@@

\subsection{Pure Domain Decomposition Scaling}
\label{subsec:pure_domain_decomp}
For the first set of scaling studies, we consider the parallel MCSA
algorithm where MSOD has not been used in the parallel Monte Carlo
algorithm. We refer to this case as pure domain decomposition where no
overlap or replication has been leveraged. Such a test will
specifically isolate the performance of Brunner and Brantley's
algorithm as applied to Neumann and Ulam's solver. For the strong
scaling study, the global diffusion problem size was fixed at a
relatively small \sn{1.6}{7} DOFs and the 16 core run used as the
reference case. Figure~\ref{fig:titan_pure_strong_time} gives the wall
time per iteration for these calculations while
Figure~\ref{fig:titan_pure_strong} gives the absolute efficiencies
computed using Eq~(\ref{eq:strong_scaling_absolute}). All MCSA
calculations converged the diffusion problem in 22 iterations.

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3.25in]{titan_pure_strong_time.pdf}
  \end{center}
  \caption{Pure domain decomposition strong scaling wall time per
    iteration.}
  \label{fig:titan_pure_strong_time}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3.25in]{titan_pure_strong.pdf}
  \end{center}
  \caption{Pure domain decomposition strong scaling absolute
    efficiency. \textit{Super-linear speed-up is from memory thrashing
      in the reference case.}}
  \label{fig:titan_pure_strong}
\end{figure}

@@

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3.25in]{titan_pure_weak_time.pdf}
  \end{center}
  \caption{Pure domain decomposition weak scaling wall time
      per iteration.}
  \label{fig:titan_pure_weak_time}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3.25in]{titan_weak_absolute.pdf}
  \end{center}
  \caption{Pure domain decomposition weak scaling absolute
    efficiency.}
  \label{fig:titan_weak_absolute}
\end{figure}

\subsection{Overlapping Domain Decomposition}
\label{subsec:overlapping_domain_decomp}
In the next set of scaling studies, we will explore how adding overlap
to the subdomains in the problem affects run times and parallel
performance as compared to the case with pure domain decomposition. To
do this, we utilize the set of analytic relations developed in
\cite{slattery_spectral_2013} that describe the distances a history in
a Neumann-Ulam random walk will travel for the neutron diffusion
problem. To determine the amount of overlap to test for the scaling
studies, we calculate how many discrete states a history will move on
average from its birth site using:
\begin{equation}
  \sqrt{\bar{n^2_k}} = n_s \sqrt{k}\:,
  \label{eq:discrete_distance}
\end{equation}
where $\sqrt{\bar{n^2_k}}$ is the root-mean-squared number of discrete
states a history will move from its birth site, $k$ the average number
of transitions a history will make in a random walk, and $n_s$ the
average number of states a history will move per transition in a
particular direction. Using \cite{slattery_spectral_2013} we compute
$k=19.22$ as the average random walk length and $n_s = 3/5$ states per
random walk step for the neutron diffusion problem giving
$\sqrt{\bar{n^2_k}} = 2.63$. This means that for every random walk, we
can expect the sequence to terminate on average 2.63 states in both
the $i$ and $j$ dimensions away from the starting state. Therefore, we
will parametrically grow the local domain using 1, 2, 3, and 5 levels
to measure how scalability is affected when this root-mean-squared
number of states is considered. For example, using 3 levels of overlap
should keep over 50\% of the histories that would be communicated in
the pure domain decomposition case on-process as this is greater than
the root-mean-squared number of states we would expect a history to
travel.

For the performance metrics, we may use the same weak and strong
scaling efficiency metrics that we used for the pure domain
decomposition case. As we are comparing to the case without overlap
and are looking to boost performance relative to this case, for both
weak and strong scaling calculations we will use the 16-core case
without overlap as the reference computation for all efficiency
measures reported. The absolute efficiency results from the strong
scaling exercise are given in
Figure~\ref{fig:titan_strong_overlap}. Aside from the boosting of
efficiencies at 2,048 cores, little evidence of strong scaling
improvement is observed for all levels of overlap tested.

@@

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3.25in]{titan_strong_overlap.pdf}
  \end{center}
  \caption{Strong scaling absolute efficiency with varying levels of
    overlap. \textit{Values of 0, 1, 2, 3, and 5 levels of overlap
      were used as the problem had a root-mean-squared number of
      states of 2.63 for each history.}}
  \label{fig:titan_strong_overlap}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=3.25in]{titan_strong_overlap_diff.pdf}
  \end{center}
  \caption{Strong scaling absolute efficiency difference with respect
    to the 0 overlap reference case. \textit{Values of 1, 2, 3, and 5
      levels of overlap are compared here with the difference of the
      absolute efficiencies presented.}}
  \label{fig:titan_strong_overlap_diff}
\end{figure}

For weak scaling, the same study as in the pure domain decomposition
case was performed again with 1, 2, 3, and 5 levels of overlap. The
absolute parallel efficiencies from these calculations are given in
Figure~\ref{fig:titan_weak_overlap} and the differences in
efficiencies from the calculations without overlap given in
Figure~\ref{fig:titan_weak_overlap_diff}. The results here are similar
in that there is not a dramatic improvement in scalability relative to
the case without overlap. Opposite the strong scaling calculation,
overlap enhances weak scaling efficiencies at lower core counts and
degrades the efficiencies at higher core counts. In addition, 1 level
of overlap provided the best enhancement of scalability when overlap
was beneficial.

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3.25in]{titan_weak_overlap.pdf}
  \end{center}
  \caption{Weak scaling absolute efficiency with varying levels of
    overlap. \textit{Values of 0, 1, 2, 3, and 5 levels of overlap
      were used as the problem had a root-mean-squared number of
      states of 2.63 for each history.}}
  \label{fig:titan_weak_overlap}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3.25in]{titan_weak_overlap_diff.pdf}
  \end{center}
  \caption{Weak scaling absolute efficiency difference with respect to
    the 0 overlap reference case. \textit{Values of 1, 2, 3, and 5
      levels of overlap are compared here with the difference of the
      absolute efficiencies presented.}}
  \label{fig:titan_weak_overlap_diff}
\end{figure}

For both the strong and weak scaling studies, the parallel performance
of the algorithm was either marginally improved or not improved at all
when overlap was added. From the literature, we would always expect an
improvement in performance when using overlap to prevent the transport
of histories from domain to domain during execution of the
Neumann-Ulam algorithm. The reduction in transport means a reduction
in parallel communication and therefore should never degrade the
performance of the algorithm. It is true that doing so in fact
enhances Monte Carlo performance. However, the Neumann-Ulam algorithm
is not representative of the entire MCSA algorithm. When overlap is
used, an overlapping tally vector is constructed in parallel such that
multiple cores may own multiple states in the system. When this
occurs, the tally data associated with these shared states must be
communicated after the Neumann-Ulam solve. This reduction operation
then builds the full MCSA correction vector in the decomposition of
the solution vector.

@@

\subsection{Multiple Set Domain Decomposition}
\label{subsec:ms_decomposition}
We next apply replication to the problem in the form of multiple
sets. As the previous analysis showed overlap to be of little
assistance in improving scalability, we only apply multiple sets to
the problem with pure domain decomposition within the sets for these
scaling studies. When adding sets, we must decide how many stochastic
histories to use to generate the correction vector in the MCSA
iteration. For this work we consider two choices; the split case and
the fully replicated case. As an example of these cases, consider a
single set problem where 20,000 histories are used in the Neumann-Ulam
sequence. If we choose the split case, then solving the same problem
with 2 sets would give 10,000 histories in each set and 4 sets would
give 5,000 histories in each set. Splitting the histories in this way
provides the same global number of histories used to compute the
correction in the single set problem and we should therefore expect
the same iterative performance at a reduced time per iteration and
thus reduced time to solution. For the replicated history case, each
additional set in the problem will use just as many histories as the
single set case. Using 2 sets with fully replicated histories for our
example problem then has each set computing 20,000 histories with
40,000 global histories in the computation. For the 4 set case, the
number of global histories is increased to 80,000. By increasing the
global number of histories, the replicated cases should approximately
maintain time per iteration but should decrease the number of
iterations required to converge and subsequently reduce time to
solution.

\subsubsection{Strong Scaling with Multiple Sets}
\label{subsubsec:ms_strong}
For the strong scaling case, recall that the objective is to decrease
the time to solution for a given global problem size $N$ by increasing
$P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref})
and (\ref{eq:algorithmic_efficiency}) to compute the various strong
scaling efficiencies, but we must modify the reference computation in
each of these formulas to reflect our objective. For all efficiency
computations, the 16 core single set computation will be used as the
reference case. By adding more cores to the problem, either by
replicating with multiple sets, or decreasing the local problem size,
any computational resources added to the problem should aim to result
in speedup of this reference case.

For the multiple sets problem, the single set strong scaling study
from the pure domain decomposition reference case was repeated for
multiple sets with both splitting of histories across sets and full
replication of histories across sets using both 2 and 4
sets. Figure~\ref{fig:titan_strong_ms_time} gives the actual wall
times for these computations. There are two important features to note
in the wall time data. First, splitting the histories among sets
instead of replicating them is faster due to the fact that cutting the
number of histories in half reduces the time of an iteration by a
factor that is nearly linear while linearly increasing the number of
histories through replication does not decrease the iterations
required to converge by a linear factor. Second, until the strong
scaling wall is hit starting at around 10,000 cores, the single set
case actually performs the fastest with only marginal improvement in
run times at higher numbers of cores with more sets. At these larger
numbers of cores, however, the improvements are so modest that one is
better off running the same problem using a single set at lower core
counts to conserve computational resources while maintaining a good
time to solution.

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=3.25in]{titan_strong_ms_time.pdf}
  \end{center}
  \caption{Wall time in seconds to solution for each case for the
    strong scaling study with multiple sets. \textit{Until the strong
      scaling wall is reached at O(10,000) cores, the single set case
      performs the best. For the multiple set cases, it is faster to
      split histories among sets and maintain iterative performance
      rather than replicate histories to improve iterative
      performance.}}
  \label{fig:titan_strong_ms_time}
\end{figure}
